---
title: "RB+AS_8803_PS2"
format: html
editor: visual
---

## ECON 8803: PS2

Rohit Borah & Abhisar Singh

February 27, 2026

```{r}
## Load requisite packages.

library(tidyverse)
library(readr)
library(gt)
library(ggthemes)
library(MatchIt)
library(glmnet)
library(hdm)
library(stargazer)
```

```{r}
## Import data.

df_raw <- read.csv(
  "/Users/bwom/Documents/PhD/Spring 2026/8803 EDA/Problem Sets/PS2/pset2.csv"
)
```

## 1. Data Cleaning

### 1a. Fix missing values.

```{r}
# Set encoded missing values to NA.
df_raw <- df_raw %>%
  mutate(
    across(
      c(
        cardiac,
        lung,
        diabetes,
        herpes,
        chyper,
        phyper,
        pre4000,
        preterm,
        tobacco,
        alcohol
      ),
      ~ na_if(., 9)
    ),
    herpes = na_if(herpes, 8),
    cigar6 = na_if(cigar6, 6),
    drink5 = na_if(drink5, 5),
    wgain = na_if(wgain, 99)
  )

# Establish set of binary variables.
binary_vars <- c(
  "anemia",
  "cardiac",
  "lung",
  "diabetes",
  "herpes",
  "chyper",
  "phyper",
  "pre4000",
  "preterm",
  "tobacco",
  "alcohol"
)

# Recode binary variables.
df_raw <- df_raw %>%
  mutate(across(all_of(binary_vars), ~ ifelse(. == 1, 1, 0)))

# str(df) # data validation test (run if needed)
```

### 1b. Produce an analysis dataset.

Produce an analysis dataset that drops any observation with missing values and verify it has 114,610 observations. Drop $\verb|stresfip|, \verb|birmon|$, and $\verb|weekday|$.

```{r}
# Remove missing values and unnecessary variables.
df <- df_raw %>%
  select(-stresfip, -birmon, -weekday) %>%
  drop_na()
```

### 1c. Produce summary statistics tables.

Describe some of the key variables in the final analysis data set. A good summary table generally includes mean, standard deviation, minimum and maximum. Pay attention to how some variables are measured (continuously, categorically, etc.).

Continuous Variables:

```{r}
# Summary Table 1: Continuous

# Establish variables
t1c_cnt_vars <- c(
  "dmage",
  "dmeduc",
  "nlbnl",
  "dlivord",
  "totord9",
  "monpre",
  "nprevist",
  "isllb10",
  "dfage",
  "dfeduc",
  "dgestat",
  "dbrwt",
  "clingest",
  "cigar6",
  "drink5",
  "wgain"
)

# Create summary statistics with variables as rows
df_1c_cnt <- data.frame(
  Variable = c(
    "Mother's Age",
    "Mother's Education",
    "Number of Live Births (Now Living)",
    "Number of Live Births (Now Dead)",
    "Total Birth Order",
    "Month Prenatal Care Began",
    "Total Number of Prenatal Visits",
    "Interval Since Last Live Birth",
    "Father's Age",
    "Father's Education",
    "Gestation - Detail in Weeks",
    "Birthweight (g)",
    "Clinical Estimate of Gestation",
    "Avg Number of Cigarettes Per Day",
    "Avg Number of Drinks",
    "Weight Gain"
  ),
  Mean = sapply(t1c_cnt_vars, function(v) {
    round(mean(df[[v]], na.rm = TRUE), 2)
  }),
  Median = sapply(t1c_cnt_vars, function(v) {
    round(median(df[[v]], na.rm = TRUE), 2)
  }),
  SD = sapply(t1c_cnt_vars, function(v) round(sd(df[[v]], na.rm = TRUE), 2)),
  Min = sapply(t1c_cnt_vars, function(v) min(df[[v]], na.rm = TRUE)),
  Max = sapply(t1c_cnt_vars, function(v) max(df[[v]], na.rm = TRUE))
)

# Format with gt
t_1c_cnt <-
  df_1c_cnt %>%
  gt() %>%
  tab_header(
    title = "Summary Statistics: Continuous Variables"
  ) %>%
  cols_label(
    Variable = "",
    Mean = "Mean",
    Median = "Median",
    SD = "SD",
    Min = "Min",
    Max = "Max"
  ) %>%
  fmt_number(
    columns = c(Mean, Median, SD),
    decimals = 2
  ) %>%
  tab_options(
    table.font.size = 11,
    heading.title.font.size = 13,
    column_labels.font.weight = "bold",
    table.width = pct(100)
  )

t_1c_cnt
```

Categorical Variables:

```{r}
# Establish variables
t1c_cat_vars <- c(
  "csex1",
  "dmar1",
  "ormoth1",
  "orfath1",
  "dplural1",
  "anemia",
  "cardiac",
  "lung",
  "diabetes",
  "herpes",
  "chyper",
  "phyper",
  "pre4000",
  "preterm",
  "tobacco",
  "alcohol"
)

# Recode categorical variables where necessary.
df <- df %>%
  mutate(dplural1 = ifelse(dplural > 1, 1, 0))
df <- df %>%
  mutate(csex1 = ifelse(csex > 1, 1, 0))
df <- df %>%
  mutate(dmar1 = ifelse(dmar > 1, 1, 0))
df <- df %>%
  mutate(ormoth1 = ifelse(ormoth %in% c(1, 2, 3, 4, 5), 1, 0))
df <- df %>%
  mutate(orfath1 = ifelse(orfath %in% c(1, 2, 3, 4, 5), 1, 0))

# Create table
df_1c_cat <- data.frame(
  Variable = c(
    "Female Children",
    "Married Mothers",
    "Hispanic Mothers",
    "Hispanic Fathers",
    "Multiple Births",
    "Anemic Mothers",
    "Mothers with Cardiac Disease",
    "Mothers with Lung Disease",
    "Diabetic Mothers",
    "Mothers with Genital Herpes",
    "Mothers with Chronic Hypertension",
    "Mothers with Pregnancy-Related Hypertension",
    "Previous Infants Weighing > 4000g",
    "Previous Preterm Infants",
    "Tobacco Use During Pregnancy",
    "Alcohol Use During Pregnancy"
  ),
  N = sapply(t1c_cat_vars, function(v) sum(!is.na(df[[v]]))),
  Proportion = sapply(t1c_cat_vars, function(v) {
    round(mean(df[[v]], na.rm = TRUE), 3)
  })
)

# Format with gt
t_1c_cat <-
  df_1c_cat %>%
  gt() %>%
  tab_header(
    title = "Summary Statistics: Categorical Variables"
  ) %>%
  cols_label(
    Variable = "Variable",
    N = "N",
    Proportion = "Proportion"
  ) %>%
  tab_options(
    table.font.size = 11,
    heading.title.font.size = 13,
    column_labels.font.weight = "bold",
    table.width = pct(100)
  )

t_1c_cat
```

### 1d. Histogram of birthweight variable.

```{r}
p_1d <- ggplot(df, aes(x = dbrwt)) +
  geom_histogram(bins = 30, fill = "#003057", color = "white") +
  geom_density() +
  labs(
    title = "1d. Histogram of Birth Weight",
    x = "Birth Weight (grams)"
  ) +
  theme_fivethirtyeight() +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.title = element_text(size = 11),
    axis.title.y = element_blank()
  )

p_1d
```

## Preliminary Analysis

### 2a. Mean ∆ Birthweight by Smoking Status

```{r}
# Build statistics table 
df_2a <- df %>%
  group_by(tobacco) %>%
  summarize(n = n(), 
            mean_brwt = mean(dbrwt), 
            sd_brwt = sd(dbrwt)) %>%
  mutate(category = ifelse(tobacco == 1, "Smoker", "Non-Smoker"), .before = n)

# Format with gt
t_2a <- df_2a %>%
  gt() %>%
  tab_header(
    title = "Birthweight by Tobacco Use"
  ) %>%
  cols_label(
    tobacco = "Group",
    category = "Category",
    n = "n",
    mean_brwt = "Mean",
    sd_brwt = "SD"
  ) %>%
  fmt_number(
    columns = c(mean_brwt, sd_brwt),
    decimals = 2
  ) %>%
  tab_options(data_row.padding.horizontal = px(10)) %>%
  cols_align(
    align = "center",
    columns = everything()
  )

t_2a
```

### 2b. Balance Table

```{r}
# Build analysis table.
df_2b <- df %>%
  group_by(tobacco) %>%
  summarise(
    n = n(),
    mean_brwt = mean(dbrwt),
    mean_age = mean(dmage, na.rm = TRUE),
    mean_education = mean(dmeduc, na.rm = TRUE),
    pct_married = mean(dmar == 1, na.rm = TRUE) * 100,
    pct_black = mean(mrace3 == 2, na.rm = TRUE) * 100, 
    mean_prenatal_visits = mean(nprevist, na.rm = TRUE),
    pct_hypertension = mean(phyper == 1, na.rm = TRUE) * 100,
    pct_diabetes = mean(diabetes == 1, na.rm = TRUE) * 100
  ) %>%
  mutate(
    tobacco_status = ifelse(tobacco == 1, "Smoker", "Non-smoker"),
    .before = n
  )

# Reformat table using pivot_longer.
df_2b_melt <- df_2b %>%
  select(-tobacco) %>%
  pivot_longer(
    cols = -tobacco_status,
    names_to = "Characteristic",
    values_to = "Value"
  ) %>%
  pivot_wider(names_from = tobacco_status, values_from = Value)

# Format output table using gt.
t_2b <- df_2b_melt %>%
  gt() %>%
  tab_header(
    title = "Balance Table by Smoking Status"
  ) %>%
  cols_label(
    Characteristic = "Characteristic",
    Smoker = "Smokers",
    `Non-smoker` = "Non-Smokers"
  ) %>%
  fmt_number(
    columns = c(Smoker, `Non-smoker`),
    decimals = 2
  ) %>%
  cols_align(
    align = "center",
    columns = c(Smoker, `Non-smoker`)
  )
t_2b
```

## 3. Regression Adjustment

### 3a. Basic model to estimate smoking impact

```{r}
# Control variables as reference
controls <- c(
  "dmage",
  "dmeduc",
  "mrace3",
  "ormoth",
  "dmar",
  "dfage",
  "dfeduc",
  "orfath",
  "nlbnl",
  "dlivord",
  "totord9",
  "isllb10",
  "monpre",
  "nprevist",
  "adequacy",
  "alcohol",
  "drink5",
  "csex"
)

# Variables to exclude as reference
excluded <- c(
  "dgestat",
  "phyper",
  "preterm",
  "pre4000",
  "anemia",
  "cardiac",
  "lung",
  "diabetes",
  "herpes",
  "chyper"
)

# Linear model
m_3a <- lm(
  dbrwt ~ tobacco +
    dmage +
    dmeduc +
    mrace3 +
    ormoth +
    dmar +
    dfage +
    dfeduc +
    orfath +
    nlbnl +
    dlivord +
    totord9 +
    isllb10 +
    monpre +
    nprevist +
    adequacy +
    alcohol +
    drink5, +
    csex,
  data = df
)

summary(m_3a)

# stargazer(m_3a)
```

### 3b. Adding in Bad Controls

```{r}
# linear model + bad controls
m_3b <- lm(
  dbrwt ~ tobacco +
    dmage +
    dmeduc +
    mrace3 +
    ormoth +
    dmar +
    dfage +
    dfeduc +
    orfath +
    nlbnl +
    dlivord +
    totord9 +
    isllb10 +
    monpre +
    nprevist +
    adequacy +
    alcohol +
    drink5 +
    csex +
    dgestat +
    phyper +
    preterm +
    pre4000,
  data = df
)

summary(m_3b)

```

### 3c. Adding in cigar6 to the 3a Model.

```{r}
# original linear model + cigar6
m_3c <- lm(
  dbrwt ~ tobacco +
    dmage +
    dmeduc +
    mrace3 +
    ormoth +
    dmar +
    dfage +
    dfeduc +
    orfath +
    nlbnl +
    dlivord +
    totord9 +
    isllb10 +
    monpre +
    nprevist +
    adequacy +
    alcohol +
    drink5 +
    csex +
    cigar6,
  data = df
)

summary(m_3c)
```

### 3d. Interacting smoking status with diabetes.

```{r}
# linear model interacting tobacco * diabetes
m_3d <- lm(
  dbrwt ~ tobacco * diabetes +
    dmage +
    dmeduc +
    mrace3 +
    ormoth +
    dmar +
    dfage +
    dfeduc +
    orfath +
    nlbnl +
    dlivord +
    totord9 +
    isllb10 +
    monpre +
    nprevist +
    adequacy +
    alcohol +
    drink5 +
    csex,
  data = df
)

summary(m_3d)
```

### 3e. Oaxaca-Blinder

```{r}
# Oaxaca-Blinder: linear model interacting tobacco with 
# all covariates

m_3e <- lm(
  dbrwt ~ tobacco *
    (dmage +
      dmeduc +
      mrace3 +
      ormoth +
      dmar +
      dfage +
      dfeduc +
      orfath +
      nlbnl +
      dlivord +
      totord9 +
      isllb10 +
      monpre +
      nprevist +
      adequacy +
      alcohol +
      drink5),
  data = df
)

# Create new tables filtered by smoking status
df_3e_smoking <- df %>%
  filter(tobacco == 1)

df_3e_nonsmoking <- df %>%
  filter(tobacco == 0)

# Predicted birthweights by smoking status
pred_smoking <- predict(m_3e, newdata = df_3e_smoking)
pred_nonsmoking <- predict(m_3e, newdata = df_3e_nonsmoking)

# ATE: ∆ means
ate_3e <- mean(pred_smoking) - mean(pred_nonsmoking) # = -240.478
```

## 4. Propensity Score Methods

### 4a. LPM to Estimate Propensity Score

```{r}
# Build linear model
m_4a_ps <- lm(
  tobacco ~ dmage +
    dmeduc +
    mrace3 +
    ormoth +
    dmar +
    dfage +
    dfeduc +
    orfath +
    nlbnl +
    dlivord +
    totord9 +
    isllb10 +
    monpre +
    nprevist +
    adequacy +
    alcohol +
    drink5 +
    csex,
  data = df
)

# Add propensity score to working dataset
df$p_score <- predict(m_4a_ps)

# Quick hist to see distribution of p-scores
hist(df$p_score) 

# OLS Regression of Birthweight on Smoking Status
# using p-score
m_4a_ols <- lm(dbrwt ~ tobacco + p_score, data = df)

summary(m_4a_ols)
```

### 4b. Logit to Estimate Propensity Score

```{r}
# Use glm to run a logit specification for p-score
m_4b_logit <- glm(
  tobacco ~ dmage +
    dmeduc +
    mrace3 +
    ormoth +
    dmar +
    dfage +
    dfeduc +
    orfath +
    nlbnl +
    dlivord +
    totord9 +
    isllb10 +
    monpre +
    nprevist +
    adequacy +
    alcohol +
    drink5 +
    csex,
  data = df,
  family = binomial(link = "logit")
)

summary(m_4b_logit)

# Add logit propensity score to working dataset
df$pscore_logit <- predict(m_4b_logit, 
                           newdata = df, 
                           type = "response")

```

### 4c. Logit Specification Overlap

```{r}
# Visualize p-score overlaps using density plots.
p_4c <-
  ggplot(df, aes(x = pscore_logit, fill = factor(tobacco))) +
  geom_density(alpha = 0.5) +
  labs(title = "4c. Propensity Score Overlap Plot", 
       fill = "Smoking Status") +
  theme_fivethirtyeight() +
  theme(legend.position = "top", 
        legend.title = element_text(face = "bold"))

p_4c
```

### 4d. ATE Using OLS Controlling for Propensity Score

```{r}
# linear model using logit pscore
m_4d <- lm(dbrwt ~ tobacco + pscore_logit, data = df)

summary(m_4d)
```

### 4e. ATE Using Blocking-On-Propensity-Score

```{r}
# Establish 10 bins by percentile
df_4e <- df %>%
  mutate(bin_pscore = ntile(pscore_logit, 10))

# Calculate ∆ mean by bin
df_4e_block <- df_4e %>%
  group_by(bin_pscore) %>%
  summarize(
    n = n(),
    mean_diff = mean(dbrwt[tobacco == 1]) - mean(dbrwt[tobacco == 0])
  )

# Blocking estimator for ATE of interest
ate_block <- sum(df_4e_block$mean_diff * df_4e_block$n / sum(df_4e_block$n))

ate_block #-226.367
```

### 4f. ATE Using Hajek (IPW)

```{r}
# Calculate Hajek IPW estimator manually
ate_treated <- sum(df$tobacco * df$dbrwt / df$pscore_logit) /
  sum(df$tobacco / df$pscore_logit)

ate_control <- sum((1 - df$tobacco) * df$dbrwt / (1 - df$pscore_logit)) /
  sum((1 - df$tobacco) / (1 - df$pscore_logit))

ate_ipw <- ate_treated - ate_control

ate_ipw # -223.348g
```

## 5. Doubly Robust Methods

### 5a. P-Score Matching with Regression Adjustment

```{r}
# Perform propensity score matching
matched <- matchit(
  tobacco ~ dmage +
    dmeduc +
    mrace3 +
    ormoth +
    dmar +
    dfage +
    dfeduc +
    orfath +
    nlbnl +
    dlivord +
    totord9 +
    isllb10 +
    monpre +
    nprevist +
    adequacy +
    alcohol +
    drink5 +
    csex,
  data = df,
  method = "nearest",
  distance = df$pscore_logit
)

# summary(matched)

# Establish matched dataset
matched_data <- match.data(matched)

# Regress on matched data
matched_reg <- lm(
  dbrwt ~ tobacco +
    dmage +
    dmeduc +
    mrace3 +
    ormoth +
    dmar +
    dfage +
    dfeduc +
    orfath +
    nlbnl +
    dlivord +
    totord9 +
    isllb10 +
    monpre +
    nprevist +
    adequacy +
    alcohol +
    drink5 +
    csex,
  data = matched_data,
  weights = weights
)

summary(matched_reg)
```

### 5b. Double-Selection LASSO

```{r}
# Create double-selection lasso.
m_5b_ds_lasso <- rlassoEffect(
  x = model.matrix(
    ~ dmage +
      I(dmage^2) +
      dmeduc +
      I(dmeduc^2) +
      mrace3 +
      ormoth +
      dmar +
      I(dmar^2) +
      dfage +
      I(dfage^2) +
      dfeduc +
      I(dfeduc^2) +
      nprevist +
      I(nprevist^2) +
      orfath +
      nlbnl +
      dlivord +
      totord9 +
      isllb10 +
      monpre +
      adequacy +
      alcohol +
      drink5 +
      csex +
      dmage * dmeduc +
      dmage * dmar +
      dmage * mrace3,
    data = df
  ),
  y = df$dbrwt,
  d = df$tobacco
)

summary(m_5b_ds_lasso)
```
