---
title: "8803_PS5_qmd"
author: "Rohit Borah"
format: html
editor: visual
---

## ECON 8803: PS5

Rohit Borah & Mingyu Li

April 23, 2026

```{r}
## Load requisite packages.

library(readr)
library(tidyverse)
library(ggthemes)
library(gt)
library(dfidx)
library(mlogit)
library(stargazer)
library(modelsummary)
```

```{r}
## Import data.
df_ps5 <- read_csv("ps5_data.csv")

# Working dataset
df <- df_ps5

## str(df)
```

### 1. Summary Statistics Table

Create a summary statistics table summarizing the characteristics of each car model chosen and not chosen across all consumers.

```{r}
## 1. Summary statistics data.

df_stats <- df %>%
  group_by(cs, choice) %>%
  summarize(
    n = n(),
    mean_price = round(mean(price), 3),
    sd_price = round(sd(price), 3),
    mean_quality = round(mean(quality), 3),
    sd_quality = round(sd(quality), 3),
    mean_cost = round(mean(cost), 3),
    sd_cost = round(sd(cost), 3),
    mean_dist_price = round(mean(dist_price), 3),
    sd_dist_price = round(sd(dist_price), 3),
    mean_dist_quality = round(mean(dist_quality), 3),
    sd_dist_quality = round(sd(dist_quality), 3),
    mean_dist_cost = round(mean(dist_cost), 3),
    sd_dist_cost = round(sd(dist_cost), 3),
    .groups = "drop"
  ) %>%
  mutate(
    Model = cs,
    Status = if_else(choice == 1, "Chosen", "Not chosen")
  ) %>%
  select(Model, Status, n,
         mean_price, sd_price,
         mean_quality, sd_quality,
         mean_cost, sd_cost,
         mean_dist_price, sd_dist_price,
         mean_dist_quality, sd_dist_quality,
         mean_dist_cost, sd_dist_cost)

# View(df_stats)
```

### 2. Conditional Logit Model

Estimate a conditional logit model for the random utility model in Equation (1). Do you think the coefficients have the expected signs? Why or why not?

```{r}
# 2. Conditional logit model

df_mlogit <- dfidx(df,
 choice = "choice",
 idx = c("id", "cs"))

m_2 <- mlogit(choice ~ price + quality + cost - 1, data = df_mlogit)

summary(m_2)
```

### 3. Distance Variables + Conditional Logit Model

Now add the distance variables in the above model. Do the coefficients of the variables change? Why do you think that is?

```{r}
# 3. Distance Variables + Conditional Logit

m_3 <- mlogit(choice ~ price + quality + cost + dist_price + dist_quality + dist_cost - 1,
 data = df_mlogit)

summary(m_3)

# stargazer(m_2, m_3)
```

### 4. Testing IIA Preferences

Next, use the distance variables to test the null hypothesis of IIA preferences. In particular, for the conditional logit model you estimated in part 3, test the null hypothesis that variables measuring characteristics of rival products are excluded (as in Equation 1).

What do you observe? (Note: Some of the distance variable are highly correlated with each other, so you may conduct this test by adding one distance variable at a time.)

```{r}
# 4. Testing IIA Preferences

# Models with one distance variable at a time
m_4a <- mlogit(choice ~ price + quality + cost + dist_price - 1, data = df_mlogit)

m_4b <- mlogit(choice ~ price + quality + cost + dist_quality - 1, data = df_mlogit)

m_4c <- mlogit(choice ~ price + quality + cost + dist_cost - 1, data = df_mlogit)

# Models with 2/3 distance variables

m_4d <- mlogit(choice ~ price + quality + cost + dist_price + dist_quality - 1, data = df_mlogit)

m_4e <- mlogit(choice ~ price + quality + cost + dist_price + dist_cost - 1, data = df_mlogit)

m_4f <- mlogit(choice ~ price + quality + cost + dist_quality + dist_cost - 1, data = df_mlogit)


# LR tests vs restricted model (m_2): H0 = distance variable excluded (IIA holds)
lrtest(m_2, m_3) # all three jointly
lrtest(m_2, m_4a) # dist_price
lrtest(m_2, m_4b) # dist_quality
lrtest(m_2, m_4c) # dist_cost
lrtest(m_2, m_4d) # dist_price + dist_quality
lrtest(m_2, m_4e) # dist_price + dist_cost
lrtest(m_2, m_4f) # dist_quality + dist_cost

# Combining tests into one list
 lr_tests <- list( 
 "dist_price" = lrtest(m_2, m_4a), 
 "dist_quality" = lrtest(m_2, m_4b), 
 "dist_cost" = lrtest(m_2, m_4c), 
 "dist_price + dist_quality" = lrtest(m_2, m_4d), 
 "dist_price + dist_cost" = lrtest(m_2, m_4e), 
 "dist_quality + dist_cost" = lrtest(m_2, m_4f), 
 "all three" = lrtest(m_2, m_3)
 ) 
 
# Combining tests into one table
t_4 <- map_dfr(lr_tests,
 ~as.data.frame(.x)[2, c("Df", "LogLik", "Chisq", "Pr(>Chisq)")],
 .id = "Variables Added") %>% 
 gt() %>% 
 tab_header(title = "LR Tests: H0 = Distance Variables Excluded (IIA)") %>%
 fmt_number(columns = Chisq, decimals = 3) %>% 
 fmt_number(columns = `Pr(>Chisq)`, decimals = 4) %>%
 cols_label(Df = "df", Chisq = "Chi-sq", `Pr(>Chisq)` = "p-value") 

t_4
```

### 5a. Effect of increasing price

Based on the demand elasticities shown above, what can you say about the effect of increasing price of the most popular product (2 lines max)?

```{r}
# 5a. Effect of increasing prices.

## No code required.
```

### 5b. Elasticity Matrix

Now, for the conditional logit model you estimated in part 2, compute the elasticity matrix and show it in the same format as the one shown above.

```{r}
# 5b. Elasticity Matrix: CLAUDE

## Note to GD: this was written almost wholly by Claude; we did not know how to do this.

# Predicted probabilities (n x J matrix) and price coefficient from m_2
probs  <- fitted(m_2, outcome = FALSE)
beta_p <- coef(m_2)["price"]

# Price matrix in wide format (n x J), sorted by individual
price_wide <- df %>%
  arrange(id, cs) %>%
  pivot_wider(id_cols = id, names_from = cs, values_from = price) %>%
  select(-id) %>%
  as.matrix()

# Compute average elasticities
J    <- 6
elas <- matrix(0, J, J)

for (j in 1:J) {
  for (k in 1:J) {
    if (j == k) {
      elas[j, k] <- mean(beta_p * price_wide[, k] * (1 - probs[, j]))
    } else {
      elas[j, k] <- mean(-beta_p * price_wide[, k] * probs[, k])
    }
  }
}

# Build display table
elas_df <- as.data.frame(elas) %>%
  setNames(as.character(1:J)) %>%
  mutate(
    Model = 1:J,
    Share = round(colMeans(probs), 3),
    Price = round(colMeans(price_wide), 3)
  ) %>%
  select(Model, Share, Price, everything())

t_5b <- elas_df %>%
  gt() %>%
  tab_header(title = "5b. Elasticity Matrix — Conditional Logit (Model 2)") %>%
  fmt_number(columns = -Model, decimals = 3) %>%
  fmt_number(columns = Model, decimals = 0)

t_5b
```

### 5c. IIA Property Across Models

Intuitively (without using any math), what can you say about the IIA property across mixed logit vs conditional logit in this context (3-4 lines max)?

```{r}
# 5c. IIA Property Across Models

## No code required.

```
