---
---
---

# Econ 8803 PS1

## Rohit Borah & Ping Yang \| February 2, 2026

### Load requisite packages.

```{r}
library(tidyverse)
library(ggthemes)
library(gt)
library(stargazer)

options(scipen = 10)
options(digits = 3)
```

### Question 1: National Supported Work Demonstration

```{r}
## Import and manipulate data.

df_jtrain2 <-
  as.data.frame(
    read_csv(
      "~/Documents/PhD/Spring 2026/8803 EDA/Problem Sets/PS1/Data/JTRAIN2.csv",
      show_col_types = FALSE
    )
  )
df_jtrain2 <- df_jtrain2 %>%
  mutate(id = row_number()) %>%
  select(id, everything())
```

##### (a) Run a regression of the outcome (earnings in 1978 in \$1000s) on the training dummy, linearly controlling for age, marriage status, and earnings in 1974. Report the coefficients and standard errors with 3 significant digits. Provide an interpretation of the coefficient of the treatment dummy.

```{r}
lm_1a <- lm(re78 ~ train + age + married + re74,
            data = df_jtrain2)

summary(lm_1a)

# stargazer(lm_1a, title = "Regression 1a", align = TRUE) # Latex table generator
```

##### (b) Verify that you get the same result for $\hat\beta$ if you first run $Y_i$ and the treatment dummy $D_i$ on other covariates, take residuals, and then run the residuals from $Y_i$ regression on the residuals from the $D_i$ regression.

```{r}
### Run regressions
lm_1by <- lm(re78 ~ age + married + re74,
             data = df_jtrain2)
lm_1bd <- lm(train ~ age + married + re74,
             data = df_jtrain2)

### Store residuals
lm_1by_residuals <- residuals(lm_1by)
lm_1bd_residuals <- residuals(lm_1bd)


### Regress y_residuals on d_residuals
lm_residuals <- lm(lm_1by_residuals ~ lm_1bd_residuals)

summary(lm_residuals)

# stargazer(lm_residuals)
```

##### (c) Provide an intuitive explanation for why you may get a similar result in (b).

##### (d) Approximate the CEF of earnings with respect to age.

i\. Estimate the CEF, $h(.) = \mathbb E[\verb|earnings|_{78} | \verb|age|]$, non-parametrically.

```{r}

### Note: we did not bin due to the number of observations per integer.
df_di <- df_jtrain2 %>%
  group_by(age) %>%
  summarize(count = n(),
            mean_re78 = mean(re78))

### Note to GD: I used AI to help write the nonparametric estimation code as
### I am unfamiliar with this.
cef <- loess(re78 ~ age,
             data = df_jtrain2)

df_jtrain2$np_cef <- fitted(cef)
```

ii\. Approximate the CEF using a linear regression and a quadratic regression of earnings on age.

```{r}

### Linear regression
lm_dii_linear <- lm(re78 ~ age,
                    data = df_jtrain2)

summary(lm_dii_linear)

df_jtrain2$lm_linear <- fitted(lm_dii_linear)

### Quadratic regression
lm_dii_quad <- lm(re78 ~ age + I(age ^ 2),
                  data = df_jtrain2)

summary(lm_dii_quad)

df_jtrain2$lm_quad <- fitted(lm_dii_quad)
```

iii\. Plot the non-parametric CEF, linear, and quadratic fits on the scatter plot between earnings and age. Clearly label which curve corresponds to each specification. What do you observe?

```{r}
p_1d <- ggplot(data = df_jtrain2,
             aes(x = age,
                 y = re78)) +
  geom_jitter() +
  geom_line(data = df_di,
              aes(x = age,
                  y = mean_re78,
                  color = "CEF"),
              se = FALSE) +
  geom_smooth(
    aes(color = "Linear"),
    method = "lm",
    formula = y ~ x,
    se = FALSE
  ) +
  geom_smooth(
    aes(color = "Quadratic"),
    method = "lm",
    formula = y ~ x + I(x ^ 2),
    se = FALSE
  ) +
  theme_fivethirtyeight() +
  labs(
    title = "8803 PS1 d.iii: Scatter Plot Between Earnings and Age",
    subtitle = "Overlaid with non-parametric CEF, linear regression, and quadratic regression",
    caption = "Rohit Borah & Ping Yang",
    x = "Age",
    y = "1978 Earnings (in $1000s)",
    color = "Specification"
  ) +
  theme(legend.position = "bottom",
        axis.title = element_text())

p_1d
```

iv\. Now, using non-parametric CEF as the proxy of true CEF, calculate the mean square error of linear and quadratic fits. Which specification provides a better approximation and why?

```{r}
mse_linear <- mean((df_jtrain2$np_cef - df_jtrain2$lm_linear) ^ 2)

mse_quad <- mean((df_jtrain2$np_cef - df_jtrain2$lm_quad) ^ 2)
```

##### (e) Now focus on the estimate of the training dummy \$\\hat\\beta\$. The estimator can be written as: $\hat\beta = \sum_i\omega_i(\textbf{X})Y_i$ where $\omega_i(\textbf{X})$ are the implicit weights placed on each outcome, which depend on the full regressor matrix **X**. Recover these weights using the following two techniques:

i\. Using FWL, derive the expression for these weights: $\omega_i = \frac{\bar D_i}{\sum_j \bar D_j^2}$

where $\bar D_i$ is the residual from regressing the treatment dummy on the controls. Plot the histogram of $\omega_i(\textbf{X})$ across all observations.

```{r}
weights_fwl <- lm_1bd_residuals / sum(lm_1bd_residuals^2)

p_1e_fwl_1 <- hist(weights_fwl, 
     main = "Histogram of Implicit Weights (FWL)",
     xlab = "Weights",
     col = "#003057",
     breaks = 40)

p_1e_fwl_1
```

ii\. Using the matrix formula $\hat\beta = (X'X)^{-1}X'Y$, compute the weights directly. Plot the histogram of $\omega_i(\textbf{X})$ across all observations. Is it similar to part (i)?

```{r}
matrix_1e_1 <- model.matrix(~ train + age + married + re74,
                               data = df_jtrain2)

matrix_1e_2 <- solve(t(matrix_1e_1) %*% matrix_1e_1) %*% t(matrix_1e_1)

weights_matrix <-matrix_1e_2[2, ]

p_1e_fwl_2 <- hist(weights_matrix,
     main = "Histogram of Implicit Weights (Matrix)",
     xlab = "Weights",
     col = "#B3A369",
     breaks = 40)

p_1e_fwl_2
```

iii\. Intuitively, what does the variation in these weights tell you about the identification of the treatment indicator?

### Question 2: Nicaragua

```{r}
## Import and manipulate data.
df_n <-
  read_csv(
    "~/Documents/PhD/Spring 2026/8803 EDA/Problem Sets/PS1/Data/Nicaragua_RCT.csv",
    ,
    show_col_types = FALSE
  )

variables <-
  c(
    "complier",
    "income_r1",
    "income_r2",
    "rubro",
    "age",
    "education",
    "mobcap",
    "fixcap",
    "land"
  )
```

##### (a) Under successful randomization, what would we expect the value of the following expression to be?

$\mathbb E[complier_i | assign_i = 1] - \mathbb E[complier_i | assign_i = 0]$

Compute the sample analogues of the above expression. Use a difference-of-means test to check if the difference is statistically significant. What do you observe?

```{r}
df_n %>%
  group_by(assign) %>%
  summarize(x_complier = mean(complier))

t_2a <- t.test(complier ~ assign, data = df_n)
t_2a
```

##### (b) Under successful randomization, what other variable(s) should be uncorrelated with $\verb|assign|$? Again, comparing sample means among these variables, is there any evidence that treated and control households are unbalanced?

```{r}
### Create df for E[var|assign = 1]
df_assign_1 <- df_n %>%
  filter(assign == 1) %>%
  summarize(
    complier = mean(complier, na.rm = TRUE),
    income_r1 = mean(income_r1, na.rm = TRUE),
    income_r2 = mean(income_r2, na.rm = TRUE),
    rubro = mean(rubro, na.rm = TRUE),
    age = mean(age, na.rm = TRUE),
    education = mean(education, na.rm = TRUE),
    mobcap = mean(mobcap, na.rm = TRUE),
    fixcap = mean(fixcap, na.rm = TRUE),
    land = mean(land, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = everything(),
               names_to = "variable",
               values_to = "mean_assign_1")

### Create df for E[var|assign = 0]
df_assign_0 <- df_n %>%
  filter(assign == 0) %>%
  summarize(
    complier = mean(complier, na.rm = TRUE),
    income_r1 = mean(income_r1, na.rm = TRUE),
    income_r2 = mean(income_r2, na.rm = TRUE),
    rubro = mean(rubro, na.rm = TRUE),
    age = mean(age, na.rm = TRUE),
    education = mean(education, na.rm = TRUE),
    mobcap = mean(mobcap, na.rm = TRUE),
    fixcap = mean(fixcap, na.rm = TRUE),
    land = mean(land, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = everything(),
               names_to = "variable",
               values_to = "mean_assign_0")

### Run t-tests for each variable.
### Note for GD- this was assisted by AI for speed.

p_values <- sapply(variables, function(v) {
  t.test(df_n[[v]][df_n$assign == 1], df_n[[v]][df_n$assign == 0])$p.value
})

### Create joined df
df_2b <- left_join(df_assign_1, df_assign_0,
                   by = "variable") %>%
  mutate(
    delta = mean_assign_1 - mean_assign_0,
    p_value = p_values,
    sig = ifelse(p_value < 0.05, "Yes", "No")
  )

df_2b %>% 
  gt() %>% 
  cols_label(
    variable = "Variable",
    mean_assign_1 = "Mean (Treated)",
    mean_assign_0 = "Mean (Control)",
    delta = "Difference",
    p_value = "P-value",
    sig = "Significant"
  ) %>% 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(columns = everything())
  ) %>% 
  fmt_number(decimals = 3)
```

##### (c) We can define the average treatment effect on the treated (ATT), that is the effect of $\verb|treat|$ on $\verb|income|$ among the treated households, using the following linear model:

$income_i = \alpha + \beta_{ATT} \times treat_i + \epsilon_i$

If we estimate this model on the full set of households using OLS, will $\hat\beta$ be an unbiased estimate of $\beta_{ATT}$? Why or why not?

```{r}

### Comparing baseline income_r1 for compliers vs non-compliers
mean(df_n$income_r1[df_n$complier == 1], na.rm = TRUE)
mean(df_n$income_r1[df_n$complier == 0], na.rm = TRUE)
```

##### (d) Now, estimate two sets of specifications for Equation 2, one without any covariates and one with the baseline variables $\verb|rubro, age, education, mobcap, fixcap, land|.$ Provide an interpretation of $\hat\beta_{ATT}$ for both cases. Do they differ? Why or why not?

```{r}
### Without covariates
lm_2di <- lm(income_r2 ~ treat,
             data = df_n)

summary(lm_2di)

# stargazer(lm_2di)

### With baseline variables
lm_2dii <-
  lm(income_r2 ~ treat + rubro + age + education + mobcap + fixcap + land,
     data = df_n)

summary(lm_2dii)

# stargazer(lm_2dii)

```
